---
title: "nis15 final project jon"
author: "Jonathan Der"
date: "2023-03-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "BIS015 PROJECT"
author: "Isaac Yang"
date: "2023-03-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load Up Library
```{r}
library(tidyverse)
library(devtools)
#devtools::install_github("dkahle/ggmap")
library(ggmap)
library(janitor)
#install.packages("ggthemes")
library(ggthemes)
```
#Exploring which theme looks beest
```{r}
ls("package:ggthemes")[grepl("theme_", ls("package:ggthemes"))]
```


#Load in Paleontology Data


```{r}
bryon_paleontology <- read_csv("C:/Users/jonat/OneDrive/Desktop/BIS15L_Paeontology/Data/bryon-paleobiodb/pbdb_data.csv")
```
#Clean the Data
```{r}
bryon_paleontology <- janitor::clean_names(bryon_paleontology)
```

#Take a look at the Data
```{r}
#mainly lots of categorical data, many bar plots
glimpse(bryon_paleontology)
```
```{r}
options(scipen = 999)
```

#Just doing counts of names to see what data to analyze
```{r}
bryon_paleontology %>% 
  count(abund_value)
```


```{r}
#Areas of interest: max_ma, min_ma, ref_pubyr, plant_organ, abund_value, state/county, the lithification ones, pres_mode, bioerosion, have up to phylum
names(bryon_paleontology)
```

#State of the Specimens
```{r}
bryon_paleontology %>%
  filter(!is.na(preservation_quality)) %>% 
  ggplot(aes(x=preservation_quality))+geom_bar()+
  theme(plot.title = element_text(size = rel(1.25), hjust = 0.5))+
  labs(title = "Specimen Quality",
      x = "Quality",
      y = "Amount of Specimens")+
  theme_clean()
```

#Construct a Bounding Box
```{r}
bryon_paleontology %>% 
   filter(!is.na(lat) & !is.na(lng)) %>% 
  select(lng, lat) %>% 
  summary()
```
```{r}
longitude_paleo <- c(-179, 179)
latitude_paleo <- c(-85, 100)
bbox_paleo <- make_bbox(longitude_paleo, latitude_paleo, f = 0.05)
bbox_paleo
```

#Create Map (Specimen Locations)
```{r}
map_paleo <- get_map(bbox_paleo, source = "stamen")
```
```{r}
ggmap(map_paleo)+
  geom_point(data = bryon_paleontology, aes(lng, lat, color = class)) +
  labs(x= "Longitude", y= "Latitude", title =" Class Locations")+
   theme(plot.title = element_text(size = rel(1.25), hjust = 0.5))
```

#Get a sense of when the data was collected
```{r}
bryon_paleontology %>% 
  ggplot(aes(x=ref_pubyr))+geom_bar(na.rm = T)+
  xlim(1800, 2020)+
  labs(title = "Paleontology Data Collection Over Time",
       x = "Year",
       y = "Specimens Found")+
   theme(plot.title = element_text(size = rel(1.25), hjust = 0.5))+
  theme_clean()
```

```{r}
#Why was there so much registered in 1999? It is nearly double that of the year in second place.
#Very hard to find why, most of the records are interested only in new discoveries
bryon_paleontology %>% 
  count(ref_pubyr, sort = T)
```

#Most Prominent Class
```{r}
#Smaller sea life makes sense, very abundant
#Mammals intersting, took over much of the world
bryon_paleontology %>%
  filter(!is.na(class)) %>%  
  count(class, sort = T) %>% 
  top_n(5, n)
```

#Bivalvia is a group of marine/freshwater molluscs that have a shell of two hinged parts nad have sideways compressed bodies.
#Gastropda are just slugs and snails
#Rhynchonellata have valves on the upper and lower with varying shell forms.
#Mammalia are mammals. They are vertebrates and produce milk.
#Cephalopoda class of mollusks prominant heads and set of tentacles.

#Look at state of how it was found
```{r}
bryon_paleontology %>%
  filter(lithology1 != "not reported") %>%  
  count(lithology1, sort = T) %>% 
  top_n(5, n)
```


```{r}
bryon_paleontology %>%
  filter(!is.na(lithology1)) %>%
  count(lithology1, sort = T) %>% 
  top_n(15, n) %>% 
  ggplot(aes(x=lithology1, y = n))+geom_col()+
  theme(plot.title = element_text(size = rel(1.25), hjust = 0.5))+
  labs(title = "Rocks Structures Surrounding Specimens (Top 15)",
      x = "Rock Structures",
      y = "Amount of Specimens")+
  theme_clean()+
  coord_flip()
```

#Focus on analyzing the most prominent ones
```{r}
#MA stands for Million Years
bryon_paleontology %>%
  filter(class == "Bivalvia" | class == "Gastropoda" | class == "Rhynchonellata" | class == "Mammalia" | class == "Cephalopoda") %>% 
  mutate(mean_ma = (max_ma+min_ma)/2) %>% 
  ggplot(aes(x = class, y = mean_ma))+geom_boxplot(na.rm = T)+
  theme(plot.title = element_text(size = rel(1.25), hjust = 0.5))+
  labs(title = "Specimen Age",
      x = "Class",
      y = "Age Range (Millions of Years)")+
  theme_clean()
```

```{r}
bryon_paleontology %>%
  filter(abund_value != "(none given)" | abund_value != "-") %>% 
  mutate(abund_value = as.character.numeric_version(abund_value)) %>% 
  filter(class == "Bivalvia" | class == "Gastropoda" | class == "Rhynchonellata" | class == "Mammalia" | class == "Cephalopoda") %>%
  ggplot(aes(x = class, y = abund_value)) + 
  geom_boxplot()+ 
  labs(title = "Distribution of Abundance of Specimens")
#how should i display the abundance values?
```

```{r}
bryon_paleontology %>% 
  count(abund_value)
```
#testing graph of diets
```{r}
bryon_paleontology %>%
  filter(class == "Bivalvia" | class == "Gastropoda" | class == "Rhynchonellata" | class == "Mammalia" | class == "Cephalopoda") %>%
  count(diet, sort = T)

#top 5 are (suspension feeder, carnivore, grazer, herbivore, omnivore)
```

```{r}
#facet wrap
bryon_paleontology%>%
  
  filter(class == "Bivalvia" | class == "Gastropoda" | class == "Rhynchonellata" | class == "Mammalia" | class == "Cephalopoda") %>%
  filter (diet == "herbivore" | diet == "omnivore" | diet == "grazer" | 
            diet == "suspension feeder" | diet == "carnivore") %>%
  ggplot(aes(x=class, fill = diet))+geom_bar()+
 # scale_y_log10()+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Diet of Top 5 Most Occuring Classes",
       x = "class",
       y = "Specimen Count")+
  theme_clean()
```
```{r}
names(bryon_paleontology)
```

#Testing Max_ma/min_ma vs abund_in_sediment (Not doing)

```{r}
bryon_paleontology%>%
  filter(class == "Bivalvia" | class == "Gastropoda" | class == "Rhynchonellata" | class == "Mammalia" | class == "Cephalopoda") 

```

# LIthology of top 5 specimen 
```{r}
lithology <- bryon_paleontology %>%
  count(lithology1, sort = T) %>% 
  top_n(15, n) 
lithology

```
```{r}
bryon_paleontology %>%
  select(class, lithology1)%>%
  filter(class == "Bivalvia" | class == "Gastropoda" | class == "Rhynchonellata" | class == "Mammalia" | class == "Cephalopoda")%>%
  filter(lithology1 == "sandstone" | lithology1 == '"limestone"' | lithology1 == '"shale"' | lithology1 == "siltstone" |
           lithology1 == "claystone" | lithology1 == '"siliciclastic"'| lithology1 == "mudstone" | lithology1 == "lime mudstone"|
           lithology1 == "not reported" | lithology1 =="marl"  | lithology1 == "packstone" | lithology1 == "wackestone" |
           lithology1 == "grainstone"| lithology1 == '"reef rocks"') %>%
  ggplot(aes(x=lithology1, fill = class))+
  #scale_y_log10()+
    geom_bar()+
  facet_grid(class~.)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Lithology by Top 5 Classes",
       x = NULL,
       y = "Specimen Count",
       fill = "Class")+
  theme_clean()
```
```{r}
living <- bryon_paleontology %>%
  filter(class == "Bivalvia" | class == "Gastropoda" | class == "Rhynchonellata" | class == "Mammalia" | class == "Cephalopoda")%>%
  count(life_habit, sort = T) %>% 
  top_n(15, n) 
living
```

```{r}
bryon_paleontology %>%
  select(class, life_habit)%>%
  filter(class == "Bivalvia" | class == "Gastropoda" | class == "Rhynchonellata" | class == "Mammalia" | class == "Cephalopoda")%>%
  ggplot(aes(x=life_habit, fill = class))+
  #scale_y_log10()+
    geom_bar()+
  facet_grid(class~.)+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title ="Life Habits of the Top 5 Classes",
       x = "Life Habit",
       y = "Specimen Count",
       fill = "Class")
```

```{r}
bryon_paleontology %>%
  select(class, life_habit)%>%
  filter(class == "Cephalopoda")%>%
  count(life_habit, sort = T)
```

